Sea-ice and macrozooplankton distribution as determinants of top predator community structure in Antarctic winter

### 2024-06-07 analysis/paper/paper.qmd
fix cross-references

We measured hydrology at each survey station. Conductivity and temperature in the water column were measured using a Sea-Bird Inc. SBE-9/11+ CTD profiler [@tbl-envvar]. We used Niskin sampling bottles to collect water samples for measuring chlorophyll *a* (chl *a*) and phaeopigment concentrations [@tbl-envvar]. For full details, see @dietrich2021winter.

@@

We assessed the community structure from relative abundance data using cluster and ordination analyses. First, we calculated a Bray-Curtis dissimilarity matrix using the R package 'vegan' [@oksanenVeganCommunityEcology2022] to estimate dissimilarities in species composition between sites. We then generated hierarchical clusters using Ward's minimum variance linkage method, implemented in the R package 'cluster' [@maechlerClusterClusterAnalysis2022], applied to the dissimilarity matrix. We determined the optimal number of clusters (*k*) to retain from the hierarchy using the gap statistic [@tibshiraniEstimatingNumberClusters2001]. The gap statistic compares the within-cluster variation of different values of *k* to their expected values under a null reference distribution. The gap statistic, *f*, usually increases with *k*, but exhibits a shoulder at the optimal number of clusters. A recommended heuristic for finding the shoulder is selecting the smallest *k* such that the gap statistic for *k*, *f*(*k*), is greater than the gap statistic for *k + 1* less the standard error, *s* [@eq-gap-stat].

@@

We used linear regression to quantify the relationship between the retained NMDS axes and the environmental, biological, and temporal variables. We performed this analysis with the `envfit()` function in the 'vegan' package, which treats environmental variables as the response variables and NMDS ordination scores as the predictors. The environmental variables included physical conditions (hydrology and ice conditions), biology (macrozooplankton community), and temporal variables [@tbl-envvar]. Temporal variables included time of day of the net tow, to account for diel patterns in macrozooplankton distribution, as well as year, to account for interannual variability. The regression coefficients obtained from the analysis represent the direction and magnitude of the maximal change in community structure associated with each environmental variable. We tested for significant associations between predator and macrozooplankton communities using a chi-squared test and assessed the statistical significance of individual predator–prey community associations via post hoc analysis of the Pearson residuals.

@@

We retained 3 predator clusters from the hierarchical analysis, which we refer to as the ‘pack ice’, ‘marginal ice’, and ‘open water’ communities based on their indicator species (Table 3) and relationships to sea-ice coverage ([@fig-seaice]). See Fig. S3 for the gap statistic curve, which we used to identify the optimal number of clusters.

@@

The open water community was observed most frequently (56.3% of sites), followed by the marginal ice (22.0%) and pack ice (21.6%) communities (Table 4), with variability between years related to the spatial distribution of sea ice ([@fig-predclustannual]). Geographically, in general, the open water community occupied the northern offshore regions and the gap between Elephant Island and the rest of the Shetland Islands ([@fig-predclustkde]). We found predators characteristic of the marginal ice community in the western Bransfield Strait and north of Elephant Island. The pack ice community primarily occupied the eastern Bransfield Strait.

@@

We retained 3 NMDS axes for investigating the relationships between the environment and community structure of predators (stress = 0.135). Two variables, sampling year and ice coverage, were significantly correlated with the ordination scores (p = 0.002 and 0.013, respectively), although the effect size was small (r^2^ = 0.061 and 0.054, respectively; Table 3; see Table S3 for the coefficient, r^2^, and p-values of the environmental fit regression). The pack ice cluster occupied a wide range of values on NMDS axis 1, overlapping with the other 2 clusters ([@fig-nmds]). The marginal ice and open water clusters were well separated by NMDS axis 2.

The predator clusters differed in temperature, salinity, and chl *a*, but not by mixed layer depth or phaeopigment (Table 5). Relative to the open water community, the pack ice community was associated with colder, more saline water with lower chl *a* concentrations. The marginal ice community occupied intermediate habitats.

The predator clusters were significantly associated with macrozooplankton clusters (χ^*2*^ test, p \< 0.001). Five pairwise associations of predators with prey were significant after applying Bonferroni correction (Table 6). The pack ice community was positively associated with macrozooplankton community 3a (an extremely diverse assemblage associated with cold, high-salinity water) and negatively associated with cluster 2b (*Thysanoessa macrura*). The marginal ice community was positively associated with macrozooplankton community 3b (large krill, including *Euphausia superba*). The open water predator community was positively associated with cluster 2b and negatively associated with 3b. Macrozooplankton cluster 1 (a ubiquitous cluster indicated by *Salpa* and *Clione*) was not significantly associated with any predator cluster. Macrozooplankton cluster 2a (indicated by important prey species such as *E. frigida*, *E. triacantha*, myctophid larvae, and *Themisto gaudichaudii*) was also not significantly associated with any predator cluster, although it was associated most often with the open water predator cluster (Table 6).

@@

We analyzed the composition and distribution of predator communities around the northern Antarctic Peninsula in winter to investigate associations among marine predators, their prey, and the physical environment. We identified 3 communities of predators, and hypothesized that the spatial organization of these communities is related to the distribution of macrozooplankton communities and a gradient in sea-ice coverage. The pagophilic community (e.g. Adélie penguins, crabeater seals) consisted of species with diets dominated by Antarctic krill (\[\@tbl-predators\]) that rely on the pack ice throughout the annual cycle [@ainleyBirdsAntarcticSea2017; @besterAntarcticMarineMammals2017]. Predators in the marginal ice zone (e.g. Antarctic fur seals, snow petrels) and in open water (e.g. southern fulmars, other petrels) communities also consume Antarctic krill, but their diets are more diverse, with a greater reliance on fish, especially myctophids such as *Electrona antarctica* (\[\@tbl-predators\]) [@ridoux1989diets; @ainley1992does; @lorentsen1998diet]. Despite the diet overlap between these 2 communities, they tend to segregate spatially according to ice coverage due to differing foraging strategies. Marginal ice zone predators use ambush and pursuit plunging behavior to exploit prey among ice floes, which open water predators have not been observed to do [@ainleySpeciesHabitatRelationshipsAntarctic1993].

@@

Species interactions play a key role in the community organization of marine predators [@stachowiczMutualismFacilitationStructure2001; @veitPositiveInteractionsForaging2017]. Identifying habitat utilization and ecological associations of individual species is crucial for establishing a robust baseline for their needs and formulating conservation strategies. However, examining habitat associations within species assemblages may provide further insights into how members of a community interact with each other [@hoffmanEcologySeabirdFeeding1981; @levinIntegratedEcosystemAssessments2009]. Further, if the presence or abundance of one species is critical to the stability and resilience of the community to environmental change, then the changes in the abundance structure of that community need to be assessed [@stachowiczMutualismFacilitationStructure2001; @dakosCriticalSlowingEarly2014]. For example, we found a predator community within the marginal ice zone that contained flying (gulls, terns, and petrels) and diving (fur seals) predators. If the aerial predators rely on diving predators to drive prey nearer to the surface [@harrisonObservationsMultispeciesSeabird1991; @veitPositiveInteractionsForaging2017], then changes in sea-ice habitat used by fur seals for haul out and rest would have additional indirect effects on the winter feeding ecology of flying seabirds within this region [@evansAssociationsSeabirdsCetaceans1982; @veitAggregationPatternsPelagic1993; @monierChangesPositiveAssociations2020]. Our analysis of species assemblages within 3 distinct communities (i.e. open water, marginal ice, and pack ice) may provide a simplified approach to examining future changes in the entire predator community, as opposed to multiple assessments of individual species’ responses to climate change. Therefore, we recommend that future studies aiming to link climate change projections, including changes in winter sea-ice concentration, assess how the 3 communities may shift or contract/expand geographically under different scenarios. Importantly, the geographic extent, environmental conditions, and macrozooplankton species associations underlying the 3 communities (Tables 4–6) are provided and can be used to develop predator community habitat models allowing for prediction in years outside of the intensive sampling period. Suggested geographic community analyses may include quantifying the habitat size and range (e.g. horizontal area and extent) of each community relative to various climate change scenarios. This community synthesis should provide helpful data for assessing climate change impacts, as well as winter-related conservation and ecosystem management needs.

>>>>

We measured hydrology at each survey station. Conductivity and temperature in the water column were measured using a Sea-Bird Inc. SBE-9/11+ CTD profiler ([@tbl-envvar]). We used Niskin sampling bottles to collect water samples for measuring chlorophyll *a* (chl *a*) and phaeopigment concentrations ([@tbl-envvar]). For full details, see @dietrich2021winter.

@@

We assessed the community structure from relative abundance data using cluster and ordination analyses. First, we calculated a Bray-Curtis dissimilarity matrix using the R package 'vegan' [@oksanenVeganCommunityEcology2022] to estimate dissimilarities in species composition between sites. We then generated hierarchical clusters using Ward's minimum variance linkage method, implemented in the R package 'cluster' [@maechlerClusterClusterAnalysis2022], applied to the dissimilarity matrix. We determined the optimal number of clusters (*k*) to retain from the hierarchy using the gap statistic [@tibshiraniEstimatingNumberClusters2001]. The gap statistic compares the within-cluster variation of different values of *k* to their expected values under a null reference distribution. The gap statistic, *f*, usually increases with *k*, but exhibits a shoulder at the optimal number of clusters. A recommended heuristic for finding the shoulder is selecting the smallest *k* such that the gap statistic for *k*, *f*(*k*), is greater than the gap statistic for *k + 1* less the standard error, *s* ([@eq-gap-stat]).

@@

We used linear regression to quantify the relationship between the retained NMDS axes and the environmental, biological, and temporal variables. We performed this analysis with the `envfit()` function in the 'vegan' package, which treats environmental variables as the response variables and NMDS ordination scores as the predictors. The environmental variables included physical conditions (hydrology and ice conditions), biology (macrozooplankton community), and temporal variables ([@tbl-envvar]). Temporal variables included time of day of the net tow, to account for diel patterns in macrozooplankton distribution, as well as year, to account for interannual variability. The regression coefficients obtained from the analysis represent the direction and magnitude of the maximal change in community structure associated with each environmental variable. We tested for significant associations between predator and macrozooplankton communities using a chi-squared test and assessed the statistical significance of individual predator–prey community associations via post hoc analysis of the Pearson residuals.

@@

We retained 3 predator clusters from the hierarchical analysis, which we refer to as the ‘pack ice’, ‘marginal ice’, and ‘open water’ communities based on their indicator species ([@tbl-indicators]) and relationships to sea-ice coverage ([@fig-seaice]). See Fig. S3 for the gap statistic curve, which we used to identify the optimal number of clusters.

@@

The open water community was observed most frequently (56.3% of sites), followed by the marginal ice (22.0%) and pack ice (21.6%) communities ([@tbl-commfreq]), with variability between years related to the spatial distribution of sea ice ([@fig-predclustannual]). Geographically, in general, the open water community occupied the northern offshore regions and the gap between Elephant Island and the rest of the Shetland Islands ([@fig-predclustkde]). We found predators characteristic of the marginal ice community in the western Bransfield Strait and north of Elephant Island. The pack ice community primarily occupied the eastern Bransfield Strait.

@@

We retained 3 NMDS axes for investigating the relationships between the environment and community structure of predators (stress = 0.135). Two variables, sampling year and ice coverage, were significantly correlated with the ordination scores (p = 0.002 and 0.013, respectively), although the effect size was small (r^2^ = 0.061 and 0.054, respectively; [@tbl-indicators]; see Table S3 for the coefficient, r^2^, and p-values of the environmental fit regression). The pack ice cluster occupied a wide range of values on NMDS axis 1, overlapping with the other 2 clusters ([@fig-nmds]). The marginal ice and open water clusters were well separated by NMDS axis 2.

The predator clusters differed in temperature, salinity, and chl *a*, but not by mixed layer depth or phaeopigment ([@tbl-clustenv]). Relative to the open water community, the pack ice community was associated with colder, more saline water with lower chl *a* concentrations. The marginal ice community occupied intermediate habitats.

The predator clusters were significantly associated with macrozooplankton clusters (χ^*2*^ test, p \< 0.001). Five pairwise associations of predators with prey were significant after applying Bonferroni correction ([@tbl-predprey]). The pack ice community was positively associated with macrozooplankton community 3a (an extremely diverse assemblage associated with cold, high-salinity water) and negatively associated with cluster 2b (*Thysanoessa macrura*). The marginal ice community was positively associated with macrozooplankton community 3b (large krill, including *Euphausia superba*). The open water predator community was positively associated with cluster 2b and negatively associated with 3b. Macrozooplankton cluster 1 (a ubiquitous cluster indicated by *Salpa* and *Clione*) was not significantly associated with any predator cluster. Macrozooplankton cluster 2a (indicated by important prey species such as *E. frigida*, *E. triacantha*, myctophid larvae, and *Themisto gaudichaudii*) was also not significantly associated with any predator cluster, although it was associated most often with the open water predator cluster ([@tbl-predprey]).

@@

We analyzed the composition and distribution of predator communities around the northern Antarctic Peninsula in winter to investigate associations among marine predators, their prey, and the physical environment. We identified 3 communities of predators, and hypothesized that the spatial organization of these communities is related to the distribution of macrozooplankton communities and a gradient in sea-ice coverage. The pagophilic community (e.g. Adélie penguins, crabeater seals) consisted of species with diets dominated by Antarctic krill ([@tbl-predators]) that rely on the pack ice throughout the annual cycle [@ainleyBirdsAntarcticSea2017; @besterAntarcticMarineMammals2017]. Predators in the marginal ice zone (e.g. Antarctic fur seals, snow petrels) and in open water (e.g. southern fulmars, other petrels) communities also consume Antarctic krill, but their diets are more diverse, with a greater reliance on fish, especially myctophids such as *Electrona antarctica* ([@tbl-predators]) [@ridoux1989diets; @ainley1992does; @lorentsen1998diet]. Despite the diet overlap between these 2 communities, they tend to segregate spatially according to ice coverage due to differing foraging strategies. Marginal ice zone predators use ambush and pursuit plunging behavior to exploit prey among ice floes, which open water predators have not been observed to do [@ainleySpeciesHabitatRelationshipsAntarctic1993].

@@

Species interactions play a key role in the community organization of marine predators [@stachowiczMutualismFacilitationStructure2001; @veitPositiveInteractionsForaging2017]. Identifying habitat utilization and ecological associations of individual species is crucial for establishing a robust baseline for their needs and formulating conservation strategies. However, examining habitat associations within species assemblages may provide further insights into how members of a community interact with each other [@hoffmanEcologySeabirdFeeding1981; @levinIntegratedEcosystemAssessments2009]. Further, if the presence or abundance of one species is critical to the stability and resilience of the community to environmental change, then the changes in the abundance structure of that community need to be assessed [@stachowiczMutualismFacilitationStructure2001; @dakosCriticalSlowingEarly2014]. For example, we found a predator community within the marginal ice zone that contained flying (gulls, terns, and petrels) and diving (fur seals) predators. If the aerial predators rely on diving predators to drive prey nearer to the surface [@harrisonObservationsMultispeciesSeabird1991; @veitPositiveInteractionsForaging2017], then changes in sea-ice habitat used by fur seals for haul out and rest would have additional indirect effects on the winter feeding ecology of flying seabirds within this region [@evansAssociationsSeabirdsCetaceans1982; @veitAggregationPatternsPelagic1993; @monierChangesPositiveAssociations2020]. Our analysis of species assemblages within 3 distinct communities (i.e. open water, marginal ice, and pack ice) may provide a simplified approach to examining future changes in the entire predator community, as opposed to multiple assessments of individual species’ responses to climate change. Therefore, we recommend that future studies aiming to link climate change projections, including changes in winter sea-ice concentration, assess how the 3 communities may shift or contract/expand geographically under different scenarios. Importantly, the geographic extent, environmental conditions, and macrozooplankton species associations underlying the 3 communities ([@tbl-commfreq; @tbl-clustenv; @tbl-predprey]) are provided and can be used to develop predator community habitat models allowing for prediction in years outside of the intensive sampling period. Suggested geographic community analyses may include quantifying the habitat size and range (e.g. horizontal area and extent) of each community relative to various climate change scenarios. This community synthesis should provide helpful data for assessing climate change impacts, as well as winter-related conservation and ecosystem management needs.
